Imprecise Datapath Design: An Overclocking Approach

### 2013-05-29 TCAD2013/overclocking_tcad.tex

\title{Imprecise Computing in Datapath Design: An Overclocking Approach}
%Imprecise Computing in Datapath Design: Precision Loss or Timing Violations}
%Imprecise Datapath Design: An Overclocking Approach
%Imprecise Computation Through Overclocking

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
As process scaling introduces significant performance variations, releasing the tight accuracy requirement rather than covering all possible worst cases would potentially offer the greater freedom to create a design with better performance or energy efficiency. In this paper, we compare two different approaches that could trade accuracy for performance. One is the traditional approach where the precision used in the datapath is limited to meet a target latency. The other is a proposed new approach to boost the clock frequency by simply allowing the datapath to operate without timing closure. We demonstrate analytically and experimentally that on average, our approach obtains either smaller errors or faster operating frequencies in comparison to the traditional approach, since the worst case caused by timing violations only happens rarely, while the precision loss results in errors to all data. We also show that for embedded applications where silicon area is also a limited factor, using the proposed approach on simple arithmetic primitives could achieve better accuracy or performance than using the traditional method on advanced operators.
\end{abstract}

@@

\section{Introduction}
\IEEEPARstart{C}{ircuit} performance has increased tremendously over the past decades, with the continuous scaling of CMOS technology to the nanometer regime. However, the drastic variations introduced by higher integration densities is anticipated to be the major obstacle when designing reliable, high performance circuits. Although hardware designers traditionally employ safety margins to ensure a uniform functionality across a variety of possible working environments, there are two major problems if we keep using this approach nowadays. One is that the uniform performance should no longer be expected in the future, because the highly scaled CMOS devices would inevitably exhibit probabilistic or statistical behavior. The other problem is that covering all possible worst cases would become increasingly difficult, expensive and result in large yield loss because of the variation.

In addition, we also notice that the clock frequency is no longer scaling significantly with time, and the dark silicon becomes an important issue in circuit design since energy efficiency is considered to be the main limitation for performance improvement in embedded or mobile applications~\cite{FutureMicroprocessor2005,DarkSilicon2011}. In this case, designers tend to employ multi-core processors, heterogeneous systems and massive parallel accelerators to sustain the performance scaling.

Specifically, large volume of current studies has demonstrated that, significant performance gains can be achieved by FPGA-based accelerators over software designs across a wide range of applications. However, one of the major factors that limits the performance of these accelerators is that they typically run at much lower clock frequencies than general purpose processors (GPPs) or GPUs. Furthermore, timing analysis tools typically report a very conservative clock frequency to guarantee ``safe'' operations. This substantially limits the potential performance of the device to a further extent.

In order to boost the operating frequency of a datapath, the standard approaches are either to heavily pipeline the design, or reduce the precision throughout the datapath. For the former method, it should be noted that pipelining will not tend to reduce the circuit latency. Actually the latency in terms of clock cycles will increase. As a result, this method will not be applicable to many embedded applications, which typically have strict latency requirements, or in datapath containing feedback where C-slow retiming is inappropriate. For the second approach, reducing the datapath precision would subtract out the latency, but at the cost of introducing quantization errors into the design. For hardware platforms such as FPGAs or ASICs, research into exploiting the potential benefits of using the minimum precision necessary to satisfy a design specification has been an extensive topic~\cite{GeorgeDT11}, as it is possible to use a number system with customized precisions on these platforms. A brief review of existing approaches in this area will be discussed in Section~\ref{Section_Background_WL}.

Unfortunately, neither of the conventional approaches tends to remove the conservative safety margin. In recent years, we have seen a growth of research that explores the potential power or performance benefits that can be obtained when relaxing the variation-induced guard-bands~\cite{NewBox2004}. The detailed review of this research field will be presented in Section~\ref{Section_Background_Imprecise}.

In this paper, we describe an alternative circuit design methodology when considering tradeoffs between accuracy, performance and area. We suggest that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work [xxx], with a further enrichment by incorporating silicon area as another evaluation metric for the trade-offs. In this work, we explicitly demonstrate the optimum design choice of the basic arithmetic operators (i.e. binary adders) under various design constraints. A summary of the main contributions of this work are as follows:

The rest of this paper is organized as follows. It first reviews in detail the current literature regarding the existing approaches that are used in word-length optimization and approximate datapath design in Section~\ref{Section_Background}. The analytical error models for two different adder structures: RCA and CSA, are discussed in Section~\ref{section_RCA} and Section~\ref{Section_CSA}, respectively. The decision of the optimum adder structure under the accuracy-performance-area trade-offs, is then put forward in Section~\ref{Section_OptimumAdder}. It is followed by a description of the probabilistic error models for CCM in Section~\ref{section_CCM}. Section~\ref{Section_ExpSetup} discusses a practical experimental setup to verify our models and test the proposed design methodology. The demonstration of the benefits of our proposed approach is then demonstrated in Section~\ref{Section_Experiments}. Finally, Section~\ref{Section_Conclusion} discusses the conclusions of this work and the possible research directions in the future.

\section{Background}\label{Section_Background}
\subsection{Word-length Optimization}\label{Section_Background_WL}
There exists a significant amount of work demonstrating that optimizing the precision used throughout the datapath would bring a substantial benefits on clock speed, silicon area and power consumption. As loosing precisions causes truncation or overflow errors, a main stream of research focus on analyzing errors generated from using the minimum precision and providing bounds of inputs to ensure the overall error are within the given accuracy specification. Current literatures report two major approaches in this area: the simulation-based approach [xxx] and the analytical-based approach such as interval arithmetic~\cite{moore1966IA}, affine arithmetic~\cite{de2004affine}, Satisfiability-Modulo Theories~\cite{SAT2010} and polynomial representations~\cite{Boland2011TCAD}. A detailed summary of these methods can be found in~\cite{GeorgeDT11, lee2006accuracy}.

\subsection{Imprecise Circuit Design Methodology}\label{Section_Background_Imprecise}
Note that, however, the choice of precision is not the only source of errors when designing a datapath. Recently we have seen a growth of parallel streams of research which aim at exploring alternative methods to trade accuracy for design efficiency. This strand of research is motivated by the drastic performance variations introduced by process scaling.

%it becomes increasingly difficult and expensive to cover all possible worst cases when designing circuits, because the rapid scaling of technology introduces drastic variations on circuit performance.

To address this problem, the international technology roadmap for semiconductors (ITRS07) pointed out that extra benefits of manufacturing, test, power and timing can be obtained if the tight requirement of absolute correctness is released for devices and interconnects~\cite{semiconductor2007international}. This topic is expected to be of growing importance in the future, because we will face new challenges as the technology scales further.

According to the existing literatures, one way to operating circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. Research in this area is based on the observations that in practice, the worst case happens rarely. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overscaling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, $22\%$ or over $30\%$ power consumption can be saved with $\sim0.01\%$ or $\sim1\%$ error rates at the output, respectively. Related work includes a similar frequency overscaling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}.

Another key observation that has been made from existing work is that the worst-case normally happens with specific input patterns. Meanwhile, many studies have shown that errors can be potentially tolerated in various applications, e.g. DSP applications concerning human perceptions. In this case, an even greater performance/power benefits can be achieved, as the design overhead of the checkers can be eliminated. Current research in this area is focusing on designing probabilistic or imprecise circuits that employ techniques from either software level or hardware level. For instance at the programming language level, a tool was proposed~\cite{EnerJ2011Uwash,Truffle2012Uwash} to divide the program into the precise parts and the approximate parts, both of which are mapped to different hardware with different speed-grades, supply voltages etc. This technique enables a relatively high service quality, in the meantime energy reduction can be obtained due to approximation. Similar ideas have also been applied on hardware directly. As an example, Palem et~al.~\cite{NonUniformScaling} described a non-uniform voltage scaling technique for the ripple carry adder. In this study, multiple voltage regions are employed for different bits along a carry chain. That is, higher voltage would be applied for computations generating most significant bits, and vice versa.

While the aforementioned literatures take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for peformance/energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et.al. described an underdesigned $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}.

However, the limitations of the existing work should not be neglected. Firstly, some techniques, such as employing several voltages regions within a ripple carry adder, are not practical to achieve in real situations, and the overhead of applying this technique is not discussed. Secondly, many studies propose alternative architectures, which means the adaptability of this kind of technique is limited on current hardware platforms. More importantly, correct results would never be obtained as long as the approximate architectures were employed. Thirdly, the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed from current research.

In this work, we propose an alternative methodology to remove these limitations. It is argued in this paper that our method can be easily applied onto existing hardware arithmetic operators such as RCAs and CCMs with almost no hardware overhead. We support this hypothesis experimentally and analytically by presenting theoretical probabilistic error models.

>>>>

\title{Imprecise Datapath Design: An Overclocking Approach}
%Imprecise Computing in Datapath Design: Precision Loss or Timing Violations}
%Imprecise Datapath Design: An Overclocking Approach --this
%Imprecise Computation Through Overclocking

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
As process scaling introduces significant performance variations, much more conservative timing margins will have to be applied to avoid timing violations. Potentially, releasing the tight accuracy requirement would offer greater freedom to create a design with better performance or energy efficiency. In this paper, we compare two different approaches that could trade accuracy for performance. One is the traditional approach where the precision used in the datapath is limited to meet a target latency. The other is a proposed new approach which simply allows the datapath to operate without timing closure. We demonstrate analytically and experimentally that on average our approach obtains either smaller errors or faster operating frequencies in comparison to the traditional approach. This is because the worst case caused by timing violations only happens rarely, while the precision loss results in errors to most data. We also show that for basic arithmetic operations such as addition, applying our approach to simple operators (i.e. ripple carry adders) can achieve better accuracy or performance than using the traditional method to both original operators and intrinsically faster operators (i.e. carry select adders) which consumes the same silicon area.

%for embedded applications where silicon area is also a limited factor, using the proposed approach on simple arithmetic primitives could achieve better accuracy or performance than using the traditional method on advanced operators.
\end{abstract}

@@

\section{Introduction}
\IEEEPARstart{C}{ircuit} performance has increased tremendously over the past decades, with the continuous scaling of CMOS technology to the nanometer regime. However, the drastic variations introduced by higher integration densities is anticipated to be the major obstacle when designing reliable, high performance circuits. Currently, hardware designers tend to employ conservative safety margins to ensure a uniform functionality across a variety of possible working environments. However, continuing with this approach could become very costly in the short term future. This is because the highly scaled CMOS devices would inevitably exhibit probabilistic or statistical behavior [xxx], meaning that we can no longer expect a uniform circuit performance. Furthermore, covering all possible worst cases would become increasingly difficult, expensive and result in large yield loss because of the variation [xxx].

%In addition, we also notice that the clock frequency is no longer scaling significantly with time, and the dark silicon becomes an important issue in circuit design since energy efficiency is considered to be the main limitation for performance improvement in embedded or mobile applications~\cite{FutureMicroprocessor2005,DarkSilicon2011}. In this case, designers tend to employ multi-core processors, heterogeneous systems and massive parallel hardware accelerators to sustain the performance scaling.

Specifically, a large volume of current studies has demonstrated that, significant performance gains can be achieved by FPGA-based accelerators over software designs across a wide range of applications [xxx]. However, one of the major factors that limits the performance of these accelerators is that they typically run at much lower clock frequencies than general purpose processors (GPPs) or graphics processing units (GPUs).

%Furthermore, timing analysis tools typically report a very conservative clock frequency to guarantee ``safe'' operations. This substantially limits the potential performance of the device to a further extent.

In order to boost the operating frequency of a datapath, the standard approaches are either to heavily pipeline the design or to reduce the precision throughout the datapath. For the former method, it should be noted that pipelining will not tend to reduce the circuit latency because the latency in terms of clock cycles will increase. As a result, this method will not be applicable to many embedded applications, which typically have strict latency requirements, or in any datapath containing feedback where C-slow retiming is inappropriate. For the second approach, reducing the datapath precision would reduce the latency at the cost of introducing quantization errors into the design. This has inspired research into exploiting the potential benefits of using the minimum precision necessary to satisfy a design specification~\cite{GeorgeDT11}. A brief review of existing approaches in this area will be discussed in Section~\ref{Section_Background_WL}.

Unfortunately, neither of these conventional approaches tends to remove the conservative safety margin. In recent years, we have seen a growth of research that explores the potential power or performance benefits that can be obtained when relaxing the variation-induced guard-bands~\cite{NewBox2004}. The detailed review of this research field will be presented in Section~\ref{Section_Background_Imprecise}.

In this paper, we describe an alternative circuit design methodology when considering tradeoffs between accuracy, performance and area. We suggest that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work [xxx], by incorporating silicon area as another evaluation metric. This enables us to find the optimum design choice of the basic arithmetic operators (i.e. binary adders) under various design constraints. A summary of the main contributions of this work are as follows:

The rest of this paper is organized as follows. It first reviews in detail the current literature regarding the approaches used in word-length optimization and approximate datapath design in Section~\ref{Section_Background}. The analytical error models for two different adder structures: RCA and CSA, are discussed in Section~\ref{section_RCA} and Section~\ref{Section_CSA}, respectively. The decision of the optimum adder structure under the accuracy-performance-area trade-offs, is then put forward in Section~\ref{Section_OptimumAdder}. It is followed by a description of the probabilistic error models for CCM in Section~\ref{section_CCM}. Section~\ref{Section_ExpSetup} discusses a practical experimental setup to verify our models and test the proposed design methodology. We then demonstration the benefits of our proposed approach in Section~\ref{Section_Experiments}. Finally, Section~\ref{Section_Conclusion} discusses the conclusions of this work and the possible research directions in the future.

\section{Background}\label{Section_Background}
\subsection{Word-length Optimization}\label{Section_Background_WL}
There exists a significant amount of work demonstrating that optimizing the precision used throughout a datapath would bring a substantial benefits on clock speed, silicon area and power consumption. There are two main research directions in this area. The first discusses methods to analyze the errors caused by the usage of finite precision and compute bounds on the worst-case errors seen at the output of the design. Related techniques include the simulation-based approach [xxx] and the analytical-based approach such as interval arithmetic~\cite{moore1966IA}, affine arithmetic~\cite{de2004affine}, Satisfiability-Modulo Theories~\cite{SAT2010} and polynomial representations~\cite{Boland2011TCAD}. A detailed summary of these methods can be found in~\cite{GeorgeDT11, lee2006accuracy}. The second stream uses these tools to choose the minimum precision within a datapath to satisfy an required accuracy specification [xxx]. In this work, we speculate and employ the maximum word-length in the datapath under a given timing constraint while ensuring no timing errors will occur. The truncation errors derived in this process are then compared to the overclocking errors generated in our proposed new scenario.

%As loosing precisions causes truncation or overflow errors, a main stream of research focus on analyzing errors generated from using the minimum precision and providing bounds of inputs to ensure the overall error are within the given accuracy specification. Current literatures report two major approaches in this area: the simulation-based approach [xxx] and the analytical-based approach such as interval arithmetic~\cite{moore1966IA}, affine arithmetic~\cite{de2004affine}, Satisfiability-Modulo Theories~\cite{SAT2010} and polynomial representations~\cite{Boland2011TCAD}. A detailed summary of these methods can be found in~\cite{GeorgeDT11, lee2006accuracy}.

\subsection{Imprecise Circuit Design Methodology}\label{Section_Background_Imprecise}
However, the choice of precision is not the only source of error when designing a datapath. Recently we have seen a growth of parallel streams of research which aim at exploring alternative methods to trade accuracy for design efficiency. This strand of research is motivated by the fact that extra benefits of manufacturing, test, power and timing can be obtained if the tight requirement of absolute correctness is released for devices and interconnects, as pointed out by the international technology roadmap for semiconductors (ITRS) in 2007~\cite{semiconductor2007international}. This topic is expected to be of growing importance in the future, because we will face new challenges as the technology scales further.

%the drastic performance variations introduced by process scaling.

%To address this problem, the international technology roadmap for semiconductors (ITRS07) pointed out that extra benefits of manufacturing, test, power and timing can be obtained if the tight requirement of absolute correctness is released for devices and interconnects~\cite{semiconductor2007international}. This topic is expected to be of growing importance in the future, because we will face new challenges as the technology scales further.

According to the existing literatures, one way to operating circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overscaling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, $22\%$ or over $30\%$ power consumption can be saved with $\sim0.01\%$ or $\sim1\%$ error rates at the output, respectively. Related work includes a similar frequency overscaling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}. However, the major problem of the BTWC work is the extra cost of silicon area due to the implementation of the checker circuits.

%Another key observation that has been made from existing work is that the worst-case normally happens with specific input patterns. Meanwhile, many studies have

Another field of study has shown that errors can be potentially tolerated in various applications, e.g. DSP applications concerning human perceptions. In this case, even greater performance/power benefits can be achieved, as the design overhead of the checkers can be eliminated. Current research in this area focuses on designing probabilistic or imprecise circuits both at a software level and a hardware level. For instance at the programming language level, a tool was proposed~\cite{EnerJ2011Uwash,Truffle2012Uwash} to divide the program into the precise parts and the approximate parts, both of which are mapped to different hardware with different speed-grades, supply voltages etc. This technique enables a relatively high service quality, in the meantime energy reduction can be obtained due to approximation. Similar ideas have also been applied on hardware directly. As an example, Palem et~al.~\cite{NonUniformScaling} described a non-uniform voltage scaling technique for the ripple carry adder. In this study, multiple voltage regions are employed for different bits along a carry chain. That is, higher voltage would be applied for computations generating most significant bits, and vice versa. Nevertheless, utilizing several voltage regions within a ripple carry adder will rarely be practical in real situations. Furthermore, the overhead of applying this technique is not addressed.

While the aforementioned literatures take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for peformance/energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et.al. described an underdesigned $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}. However, we should note that using imprecise architectures means correct results would never be obtained for certain input patterns. Another limitation is that the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed from current research. In addition, these techniques cannot be directly applied onto existing hardware platforms such as FPGAs.

%However, the limitations of the existing work should not be neglected. Firstly, some techniques, such as employing several voltages regions within a ripple carry adder, are not practical to achieve in real situations, and the overhead of applying this technique is not discussed. Secondly, many studies propose alternative architectures, which means the adaptability of this kind of technique is limited on current hardware platforms. More importantly, correct results would never be obtained as long as the approximate architectures were employed. Thirdly, the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed from current research.

In this work, we propose an alternative methodology to remove the limitations of those existing approaches. It is argued in this paper that our method can be easily applied onto existing hardware arithmetic operators such as RCAs and CCMs with almost no hardware overhead. We support this hypothesis experimentally and analytically by presenting theoretical probabilistic error models.
